Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
Update on Overleaf.

% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.
%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{cortical songs}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out of about 900, and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
Such precise structural information should therefore be included in decoding algorithm for a better understanding of neural activity.

\begin{figure}
\centering
%\includegraphics{figures/Ikegaya2004zse0150424620001.jpeg}
\caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
\end{figure}

It is interesting to make a parallel with the ``Rapid Formation of Robust Auditory Memories'' reported in~\citep{agus_rapid_2010} which uses noise patterns. They ``used random waveforms to probe the formation of new memories for arbitrary complex sounds. A behavioral measure was designed, based on the detection of repetitions embedded in noises up to $4 \si{\second}$ long''. The task is to detect the repetition of the same (frozen) noise within a trial. " Unbeknownst to listeners, some noise samples reoccurred randomly throughout an experimental block." They showed that the ``repeated exposure induced learning for otherwise totally unpredictable and meaningless sounds'' by showing that the sensitivity increases in that case. Note that ``acoustical analyses failed to reveal any obvious differences between good and bad noises'' and that ``time reversal had no significant effect on the RefRN advantage'' (which is quite surprising). The mearning is unsupervised (statistical, automatic), fast-acting (phase transition, ``insight''), and long-lasting (memorization).

TODO: re-read~\citep{luczak_packet-based_2015} Luczak A, McNaughton BL, Harris KD. Packet-based communication in the cortex. Nat Rev Neurosci. 2015;16(12):745--55.

Such observations suggest that neural groups or ensembles, rather than individual neurons, are emergent functional units of cortical activity~\citep{miller_visual_2014}. This work shows, using two-photon calcium imaging of populations of neurons from the primary visual cortex of awake mice during visual stimulation and spontaneous activity, that whereas intrinsic ensembles recur at random time intervals, visually evoked ensembles are time-locked to stimuli. It proposes that visual stimuli recruit endogenously generated ensembles to represent visual attributes. Note that evoked ensembles in response to a natural movie played in a loop were precisely timed across repetitions (Fig. 7).

% \subsubsection{Neural Variability and Sampling-Based Probabilistic Representations in the Visual Cortex~\citep{orban_neural_2016}}
% \begin{itemize}
%  \item
%   Stochastic sampling links perceptual uncertainty to neural response variability
% \item
%   Model accounts for independent changes in strength and variability of responses
% \item
%   Model predicts relationship between noise, signal, and spontaneous correlations
% \item
%   Stimulus statistics dependence of response statistics is explained
% \end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spiking motifs in neuromorphic engineering}
\label{sec:neuromorphic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy2018toward}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical Von Neumann architectures is the localized memory of this new type of chips. It is materialized by the physical connections between the processing units that can store information, an example directly inspired by synaptic plasticity of such connection is the memristor (ref). Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges (zenke2021visualizing). Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware is described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field.

@@

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spiking neural models with heterogeneous delays}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, we develop a model for the efficient detection of such PGs based on the inversion of a probabilistic model defining the generation of the raster plot as a combination of such groups. We show that such an inference can be achieved by a neural-like computation that could itself be used as a spiking neuron, as can be implemented in a neuromorphic chip for instance. A first result is to show the efficiency of such a scheme in detecting different PGs occurring at specific times in synthetic data. The representational capacity of the PGs is particularly interesting compared to traditional models of neuronal encoding using spiking frequency. Our second result is to propose a novel learning method for learning PGs in raster plots in a self-supervised manner. Finally, we demonstrate the use of this algorithm to the output of an event-based camera and how this may separate independent components from the stream of events. This end-to-end event-based computational brick could help improve the performance of current Spiking Neural Network solution currently used in neuromorphic chips.

>>>>

% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{cortical songs}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out of about 900, and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.6\linewidth]{figures/Ikegaya2004zse0150424620001.jpeg}%pcbi.1006283.g001.PNG_L.png}
\caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
\end{figure}
%----------------------------%

That results suggest that polychronous groups are not necessarily grouped on a topography and can be forming apparently randomly arranged connections. It is interesting to make a parallel with the ``Rapid Formation of Robust Auditory Memories'' reported in~\citep{agus_rapid_2010} which uses noise patterns to observe if listeners could learn to detect repeated occurrences of some (frozen) noise patterns. In paerticular, they used random waveforms to probe the formation of new memories for arbitrary complex sounds. A behavioral measure was designed, based on the detection of repetitions embedded in noises up to $4 \si{\second}$ long. The task is to detect the repetition of the same (frozen) noise within a trial. Unbeknownst to listeners, some noise samples reoccurred randomly throughout an experimental block. They showed that the ``repeated exposure induced learning for otherwise totally unpredictable and meaningless sounds'' by showing that the sensitivity increases in that case. Note also that ``acoustical analyses failed to reveal any obvious differences between good and bad noises'' and that ``time reversal had no significant effect on the detection accuracy (which is quite surprising). The learning is unsupervised (statistical, automatic), fast-acting (phase transition, ``insight''), and long-lasting (memorization).

Such observations may suggest that neural groups or ensembles, rather than individual neurons, are emergent functional units of cortical activity. ~\citep{miller_visual_2014} shows, using two-photon calcium imaging of populations of neurons from the primary visual cortex of awake mice during visual stimulation and spontaneous activity, that whereas intrinsic ensembles recur at random time intervals, visually evoked ensembles are time-locked to stimuli. It proposes that visual stimuli recruit endogenously generated ensembles to represent visual attributes. Note that evoked ensembles in response to a natural movie played in a loop were precisely timed across repetitions. % (Fig. 7).
As a summary, it seems imperative to include such precise structural information should therefore be included in decoding algorithm for a better understanding of neural activity.

% TODO: re-read~\citep{luczak_packet-based_2015} Luczak A, McNaughton BL, Harris KD. Packet-based communication in the cortex. Nat Rev Neurosci. 2015;16(12):745--55.
% \subsubsection{Neural Variability and Sampling-Based Probabilistic Representations in the Visual Cortex~\citep{orban_neural_2016}}
% \begin{itemize}
%  \item
%   Stochastic sampling links perceptual uncertainty to neural response variability
% \item
%   Model accounts for independent changes in strength and variability of responses
% \item
%   Model predicts relationship between noise, signal, and spontaneous correlations
% \item
%   Stimulus statistics dependence of response statistics is explained
% \end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spiking motifs in neuromorphic engineering}
\label{sec:neuromorphic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy2018toward}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical Von Neumann architectures is the localized memory of this new type of chips. It is materialized by the physical connections between the processing units that can store information, an example directly inspired by synaptic plasticity of such connection is the memristor (ref). Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges (zenke2021visualizing). Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware is described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field.

@@

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{sec:discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spiking neural models with heterogeneous delays}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %: >>>>>>>> Laurent is here <<<<<<<<<<<<<<

Here, we develop a model for the efficient detection of such PGs based on the inversion of a probabilistic model defining the generation of the raster plot as a combination of such groups. We show that such an inference can be achieved by a neural-like computation that could itself be used as a spiking neuron, as can be implemented in a neuromorphic chip for instance. A first result is to show the efficiency of such a scheme in detecting different PGs occurring at specific times in synthetic data. The representational capacity of the PGs is particularly interesting compared to traditional models of neuronal encoding using spiking frequency. Our second result is to propose a novel learning method for learning PGs in raster plots in a self-supervised manner. Finally, we demonstrate the use of this algorithm to the output of an event-based camera and how this may separate independent components from the stream of events. This end-to-end event-based computational brick could help improve the performance of current Spiking Neural Network solution currently used in neuromorphic chips.
